Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-12-11 paper/paper_acm.tex
reduce space v2

In what follows, we work in the context of homotopy type theory, seen
as Martin-Löf type thepry, with univalence axiom (and thus functional
extensionality) and higher inductive types.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
\begin{mymath}
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
\end{mymath} 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
\begin{align*}
  \chi &: \left(\sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n\
\fib{f}{b}\right) \\ &\qquad\xrightarrow{\sim} 
 (B \to \Type_n)
\end{align*}
%
 such that the diagram
\begin{mymath}
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
\end{mymath}
is a pullback for any $f$ with
 $n$-truncated homotopy fibers where $\Type_n^\bullet \defeq
 \sum_{A:\Type_n} A$ is the universe of pointed
$n$-truncated types and 
\begin{mymath}t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).\end{mymath}

@@

A colimit of a diagram $D$ over a graph $G$ is given by a type $P$
that defines a cocone of $D$, plus the universal property that for any
type $X$, the canonical map that transforms a function $f : P
\rightarrow X$ to a cocone of $D$ on $X$ is an isomorphism.
% 
\begin{defi}
Let $G$ be a graph, and $D$ be a diagram on $G$. 
Let $P:\Type$ together with
\begin{itemize}
\item a map $q_i : D_0(i) \to P$ for any
vertex $i:G_0$, \ie \begin{mymath}q : \prod_{i:G_0} D_0(i) \to P\end{mymath}
\item for any vertices $i,j:G_0$ and all edges $\phi:G_1(i,j)$, a path
  $p_{i,j}^\phi : q_j \circ D_1(\phi) = q_i$, \ie
  \begin{mymath}p : \prod_{i,j:G_0} \prod_{\phi:G_1(i,j)} q_j \circ D_1(\phi) = q_i.\end{mymath}
\end{itemize}

@@

% Colimits behave well with truncations, in the following sense:

% \begin{lem}
%   Let $D$ be a diagram, $m$ a truncation index, and
%   $P:\Type_{m}$. Let's call $P$ a $m$-colimit of $D$ if
%   for any other $X:\Type_m$, the
%   map
%   \begin{mymath}\lambda f:P \to X, \left( \lambda i,~f \circ q_i\, ;\, \lambda i\, j\,
%   \phi,\ \precompose_f (p_{i, j}^\phi)) \right)\end{mymath}
%   is an equivalence.

%   Then, if $\|D\|_m$ is the same diagram as $D$, but where every type
%   is $m$-truncated, $P$ is a $m$-colimit of $\|D\|_m$.
% \end{lem}

@@

This colimit $\colim D$ is given by
%
\begin{itemize}
\item a function $q:\prod_{i:G_0} D_0(i) \to \colim D$
\item for any $i,j:G_0$ and $\phi:G_1(i,j)$, a path
  $q_j \circ D_1(\phi) = q_i$.
\end{itemize}
%
Although we have proved in our formalization that $\colim D$ is
actually a colimit over $D$, we do not detail the proof here as the
existence of colimits is not used in the definition of the
sheafification process. We only make use of the following special case
of Giraud-Rezk-Lurie axioms.

@@

Section~\ref{sec:definition} presents the definition and basic
properties of left exact modalities. 
%
Section~\ref{sec:new-type-theory} discusses how a left exact modality
induces a new homotopy theory, and in which case this theory is
consistent.
%
Finally, Section~\ref{sec:examples-left-exact} introduces some
examples, and in particular the modality for the dense topology.

@@

As the definition of sheafification is based on subobject classifiers
and thus on homotopy fibers, we need to take a closer look at the way left
exact modalities preserve homotopy fibers.
%
\begin{prop}
\label{sec:defin-basic-prop}
For any $n$-truncated types $X$ and $Y$,
and any map $f:X \to Y$, the modalisation of fiber of $f$ above any element $y:Y$
is the fiber of $\modal f$ above $\eta_Y y$:
\begin{mymath}\modal \left(\sum_{x:X}  (f x = y)\right) = \sum_{x:\modal X}
(\modal f x = \eta_Y y),\end{mymath}
where $\modal f:\modal X \to \modal Y$ is defined via the equivalence {\it (iv)} and $\eta_Y$.

Moreover the following diagram commute
\begin{mymath}\xymatrix{
  \sum_{x:X} (f x = y) \ar[r]^\eta \ar[d]_\gamma & \modal \left(\sum_{x:X}  (f x = y)\right) \ar@{=}[dl]\\
  \sum_{x:\modal X} (\modal f x = \eta_Y y) & }\end{mymath}
where $\pi_1 \circ \gamma = \eta_X \circ \pi_1$, and
$\pi_2 \circ \gamma$ is given by the modalisation of
paths.
\end{prop}
\begin{proof}
% \nt{ça a du sens de mettre une preuve, non ? c'est non trivial et
%   nouveau par rapport aux travaux sur les modalités. Ou alors c'est
%   la preuve de Shulman qui tu n'as pas faite. En tout cas, il faut
%   clarifier.}

The proof is based on the following central result.
\begin{lem}
  Let $X:\Type_p$, $Y:\Type_p^\modal$ and $f:X\to Y$. If for all $y:Y$,
  $\modal (\fib f y)$ is contractible, then $\modal X = Y$.
\end{lem}
%
It is straighforward to define a map
\begin{mymath}\phi:\sum_{x:X}  (f x = y)\to
\sum_{x:\modal X} (\modal f x = \eta_Y y),\end{mymath}
using $\eta$ functions.
We just need to check that every $\modal$-fiber $\modal(\fib \phi {x;p})$ is
contractible.
Technical transformations allow one to prove
\begin{mymath}\fib\phi{x;p} = \fib s {y;p^{-1}}\end{mymath}
for some $a:\modal X$, $b:\modal Y$ and 
\begin{mymath}s:\fib{\eta_X}a \to \fib{\eta_Y}b.\end{mymath}
%
From definition of left exactness, one can deduce the following:
\begin{lem}
  Let $A,B:\Type_p$. Let $f:A\to B$. If $\modal A$ and $\modal B$ are
  contractible, then so is $\fib f b$ for any $b:B$.
\end{lem}
Thus, we just need to prove that $\modal(\fib {\eta_X} a)$ and
$\modal(\fib {\eta_Y} b)$ are contractible.
But we have the lemma:
\begin{lem}
  Let $A:\Type_p$. Then for any $y:\modal A$, $\modal(\fib {\eta_A}
  y)$ is contractible.
\end{lem}

Finally, $\modal(\fib s{y;p^{-1}})$ is contractible, so $\modal(\fib \phi {x;p})$ also, and the result is proved.
%\kq{Finish that}

%\nt{Indeed, I can't follow the proof for the moment}
\end{proof}

In the same way, left exact modalities preserve homotopy types.
\begin{prop}
  Let $k \leq n$.
  If $P:\Type_k$, then $\modal \widehat P : \Type_k$, where $\widehat P$
  is $P$ seen as a $n$-type.
\end{prop}
\begin{proof}
  An $n$-truncated type $P$ can equivalently be described as a type for
  which the unique map to $\one$ is with $n$-truncated fibers. Thus, the
  property is a direct corollary of
  Proposition~\ref{sec:defin-basic-prop} and the fact that $\modal \one =
  \one$.
\end{proof}

@@

It remains to show that $\modal_{n+1}$ is left exact and is compatible
with $\modal_{-1}$. To do that, we need to extend the notion of
compatibility and show that actually every modality $\modal_{n+1}$ is
compatible with $\modal_n$ on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1} \widehat T = \modal_n T$, where $\widehat T$ is $T$ seen as a
  $\Type_{n+1}$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{align*}
      \modal_{0} \widehat T &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right\| \\
      &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{align*}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    \begin{mymath}\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))\end{mymath}
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    \begin{mymath}\psi (u;x) = \eta_T a.\end{mymath}
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then
    \begin{align*}
      \modal_{n+2} \widehat  T &= \hspace{-1em} \sum_{u:T \to \Type_{n+1}^\modal}
                       \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n+1} (a=t)) = u\right\| \\
                     &= \hspace{-1em} \sum_{u:T \to
                       \Type_{n+1}^\modal} 
\hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n} (a=t)) = u\right\|
    \end{align*}
    because $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t)$ is in $\Type_n$.

@@

\begin{exm}
  If $\modal_{-1}$ is the double negation modality on $\HProp$ and
  $i=-1$, for any type $X$:
  \begin{align*}
  \modal X &\defeq \lnot\lnot \|X\|  \\
           &= \lnot\lnot X
  \end{align*}
  by univalence for h-propositions.

 Thus, $\modal$ is exactly the double negation modality on $\Type$
  described in~\ref{sec:examples-left-exact}.
\end{exm}

@@

Unfortunately, one lose canonicity when working in $\Type^\modal$;
indeed, for example, the term 
\begin{mymath}\left( \lambda P:\HProp,~\left\{\begin{array}{l} 0 \text{ if $P$ is
                                    false} \\ 1 \text{ if $P$ is
                                    true} \end{array}\right. \right)
                            P~:~\nat\end{mymath}
will not reduce to a numeral if the chosen property $P$ is not
decidable. However, we keep canonicity for ``old'' terms, i.e. the
ones constructed without using excluded middle.

>>>>

In what follows, we work in the context of homotopy type theory, seen
as Martin-Löf type theory, with univalence axiom (and thus functional
extensionality) and higher inductive types.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
\begin{mymath}
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
\end{mymath} 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
\begin{mymath}[0em]
  \chi : \sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n\
\fib{f}{b} \xrightarrow{\sim} 
 (B \to \Type_n)
\end{mymath}
%
 such that the diagram
\begin{mymath}
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
\end{mymath}
is a pullback for any $f$ with
 $n$-truncated homotopy fibers where $\Type_n^\bullet \defeq
 \sum_{A:\Type_n} A$ is the universe of pointed
$n$-truncated types and 
\begin{mymath}t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).\end{mymath}

@@

A colimit of a diagram $D$ over a graph $G$ is given by a type $P$
that defines a cocone of $D$, plus the universal property that for any
type $X$, the canonical map that transforms a function $f : P
\rightarrow X$ to a cocone of $D$ on $X$ is an isomorphism.
% 
\begin{defi}\label{def:colimit}
Let $G$ be a graph, and $D$ be a diagram on $G$. 
Let $P:\Type$ together with
\begin{itemize}
\item a map $q_i : D_0(i) \to P$ for any
vertex $i:G_0$, \ie \begin{mymath}[0em]q : \prod_{i:G_0} D_0(i) \to P\end{mymath}
\item for any vertices $i,j:G_0$ and all edges $\phi:G_1(i,j)$, a path
  $p_{i,j}^\phi : q_j \circ D_1(\phi) = q_i$, \ie
  \begin{mymath}[0em]p : \prod_{i,j:G_0} \prod_{\phi:G_1(i,j)} q_j \circ D_1(\phi) = q_i.\end{mymath}
\end{itemize}

@@

In~\ref{ssec:from-type-separated}, we will need to know how colimits
behave with respect to truncations. An answer is given by the following:

\begin{lem}
  Let $D$ be a diagram, $m$ a truncation index, and
  $P:\Type_{m}$. 
  Then, if $\|D\|_m$ is the same diagram as $D$, but where every type
  is $m$-truncated, $P$ is a $m$-colimit\footnote{$P$ is a $m$-colimit
    if $P$ satisfies the same property as in~\ref{def:colimit} when
    we replace $\Type$ by $\Type_m$} of $\|D\|_m$.
\end{lem}

@@

% This colimit $\colim D$ is given by
% %
% \begin{itemize}
% \item a function $q:\prod_{i:G_0} D_0(i) \to \colim D$
% \item for any $i,j:G_0$ and $\phi:G_1(i,j)$, a path
%   $q_j \circ D_1(\phi) = q_i$.
% \end{itemize}
% %
% Although we have proved in our formalization that $\colim D$ is
% actually a colimit over $D$, we do not detail the proof here as the
% existence of colimits is not used in the definition of the
% sheafification process. We only make use of the following special case
% of Giraud-Rezk-Lurie axioms.

@@

% Section~\ref{sec:definition} presents the definition and basic
% properties of left exact modalities. 
% %
% Section~\ref{sec:new-type-theory} discusses how a left exact modality
% induces a new homotopy theory, and in which case this theory is
% consistent.
% %
% Finally, Section~\ref{sec:examples-left-exact} introduces some
% examples, and in particular the modality for the dense topology.

@@

As the definition of sheafification is based on subobject classifiers
and thus on homotopy fibers, we need to take a closer look at the way left
exact modalities preserve homotopy fibers. The following is a part of
the library~\cite{hottlib}.
%
\begin{prop}
\label{sec:defin-basic-prop}
For any $n$-truncated types $X$ and $Y$,
and any map $f:X \to Y$, the modalisation of fiber of $f$ above any element $y:Y$
is the fiber of $\modal f$ above $\eta_Y y$:
\begin{mymath}[-0.3em]\modal \left(\sum_{x:X}  (f x = y)\right) = \sum_{x:\modal X}
(\modal f x = \eta_Y y),\end{mymath}
where $\modal f:\modal X \to \modal Y$ is defined via the equivalence {\it (iv)} and $\eta_Y$.

Moreover the following diagram commute
\begin{mymath}[-0.3em]\xymatrix{
  \sum_{x:X} (f x = y) \ar[r]^\eta \ar[d]_\gamma & \modal \left(\sum_{x:X}  (f x = y)\right) \ar@{=}[dl]\\
  \sum_{x:\modal X} (\modal f x = \eta_Y y) & }\end{mymath}
where $\pi_1 \circ \gamma = \eta_X \circ \pi_1$, and
$\pi_2 \circ \gamma$ is given by the modalisation of
paths.
\end{prop}
% \begin{proof}
% % \nt{ça a du sens de mettre une preuve, non ? c'est non trivial et
% %   nouveau par rapport aux travaux sur les modalités. Ou alors c'est
% %   la preuve de Shulman qui tu n'as pas faite. En tout cas, il faut
% %   clarifier.}

% The proof is based on the following central result.
% \begin{lem}
%   Let $X:\Type_p$, $Y:\Type_p^\modal$ and $f:X\to Y$. If for all $y:Y$,
%   $\modal (\fib f y)$ is contractible, then $\modal X = Y$.
% \end{lem}
% %
% It is straighforward to define a map
% \begin{mymath}\phi:\sum_{x:X}  (f x = y)\to
% \sum_{x:\modal X} (\modal f x = \eta_Y y),\end{mymath}
% using $\eta$ functions.
% We just need to check that every $\modal$-fiber $\modal(\fib \phi {x;p})$ is
% contractible.
% Technical transformations allow one to prove
% \begin{mymath}\fib\phi{x;p} = \fib s {y;p^{-1}}\end{mymath}
% for some $a:\modal X$, $b:\modal Y$ and 
% \begin{mymath}s:\fib{\eta_X}a \to \fib{\eta_Y}b.\end{mymath}
% %
% From definition of left exactness, one can deduce the following:
% \begin{lem}
%   Let $A,B:\Type_p$. Let $f:A\to B$. If $\modal A$ and $\modal B$ are
%   contractible, then so is $\fib f b$ for any $b:B$.
% \end{lem}
% Thus, we just need to prove that $\modal(\fib {\eta_X} a)$ and
% $\modal(\fib {\eta_Y} b)$ are contractible.
% But we have the lemma:
% \begin{lem}
%   Let $A:\Type_p$. Then for any $y:\modal A$, $\modal(\fib {\eta_A}
%   y)$ is contractible.
% \end{lem}

% Finally, $\modal(\fib s{y;p^{-1}})$ is contractible, so $\modal(\fib \phi {x;p})$ also, and the result is proved.
% %\kq{Finish that}

% %\nt{Indeed, I can't follow the proof for the moment}
% \end{proof}

\kq{Is the following usefull?}
In the same way, left exact modalities preserve homotopy types.
\begin{prop}
  Let $k \leq n$.
  If $P:\Type_k$, then $\modal \widehat P : \Type_k$, where $\widehat P$
  is $P$ seen as a $n$-type.
\end{prop}
\begin{proof}
  An $n$-truncated type $P$ can equivalently be described as a type for
  which the unique map to $\one$ is with $n$-truncated fibers. Thus, the
  property is a direct corollary of
  Proposition~\ref{sec:defin-basic-prop} and the fact that $\modal \one =
  \one$.
\end{proof}

@@

It remains to show that $\modal_{n+1}$ is left exact and is compatible
with $\modal_{-1}$. To do that, we need to extend the notion of
compatibility and show that actually every modality $\modal_{n+1}$ is
compatible with $\modal_n$ on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1} \widehat T = \modal_n T$, where $\widehat T$ is $T$ seen as a
  $\Type_{n+1}$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{mymath}
      \modal_{0} \widehat T = \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{mymath}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    \begin{mymath}\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))\end{mymath}
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    \begin{mymath}\psi (u;x) = \eta_T a.\end{mymath}
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then, as $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t)$ is in $\Type_n$,
    \begin{mymath}
      \modal_{n+2} \widehat = \hspace{-1em} \sum_{u:T \to
        \Type_{n+1}^\modal} 
      \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
        (\lambda t,~\modal_{n} (a=t)) = u\right\|.
    \end{mymath}

@@

If $\modal_{-1}$ is the double negation modality on $\HProp$ and
$i=-1$, $\modal$ is exactly the double negation modality on $\Type$
described in~\ref{sec:examples-left-exact}.

@@

Unfortunately, one lose canonicity when working in $\Type^\modal$;
indeed, for example, the function 
\begin{mymath}\lambda P:\HProp,~\left\{\begin{array}{l} 0 \text{ if $P$ is
                                    false} \\ 1 \text{ if $P$ is
                                    true} \end{array}\right.~:~\HProp
                              \to \nat\end{mymath}
will not reduce to a numeral when applied to a undecidable
property. However, we keep canonicity for ``old'' terms, i.e. the
ones constructed without using excluded middle.
